---
title: "Class 11"
author: "Mady Welch"
format: pdf
---

## Section 4: Population Scale Analysis

>Q13. Read the file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

```{r}
txtfile <- "https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt"
datatable <- read.table(txtfile)
head(datatable)
```

```{r}
summary(datatable)
table(datatable$geno)
```

**Genome Sample Sizes:**

**A/A = 108**

**A/G = 233** 

**G/G = 121**



```{r}
library(ggplot2)

dataplot <- ggplot(datatable) +
  aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE)

dataplot
```

**Median Expression Levels:**

**A/A ~ 32** 

**A/G ~ 25** 

**G/G ~ 20** 


>Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

The median expression level for A/A is much higher than G/G, but the ranges for A/A and G/G overlap. We can most likely assume that the SNP does slightly affect the expression of ORMDL3.








































