---
title: "Class 05: Data Visualization with GGPLOT"
author: "Mady Welch"
format: pdf
---

# Our First ggplot

To use the ggplot2 package first need to have it installed on my computer.

To install package we use the `install.packages()` command.

Can't use it until we use `library(ggplot2)`

```{r}
library(ggplot2)
ggplot()
```

```{r}
mpg
```

Our first plot of displ vs. hwy All ggplot() graphs are made in the same way.

-   data + aes + geoms

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  geom_point()
```

Can add more layers:

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  geom_point() +
  geom_line()
```

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)
```

# Plot of Gene Expression Data

First read data from online and store it in environment.

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q. How many genes?

```{r}
nrow(genes)
```

> Q. What are the columns?

```{r}
colnames(genes)
```

> Q. How many upregulated genes? What fraction of total genes is upregulated in this dataset (2 sigfigs)?

```{r}
table(genes[, "State"])
round(table(genes[, "State"])/nrow(genes)*100, 2)
```

## A first version plot of this data Condition1 vs Condition2

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2) + 
  geom_point() 
```

**Add Color:**

```{r}
p <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point()
p
```

**Change colors:**

```{r}
p + scale_colour_manual(values=c("blue", "gray", "red")) 
```

**Add plot annotations using `labs()` function:**

```{r}
p + scale_colour_manual(values=c("blue", "gray", "red")) + 
  labs(title="Gene Expression Changes Upon Drug Treatment", x="Control (no drug)", y="Drug Treatment")
```
