---
title: "Halloween Mini-Project"
author: "Mady Welch"
format: pdf
---

## Importing Candy Data

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
read.csv(candy_file)
```

```{r}
candy <-  read.csv(candy_file, row.names=1)
head(candy)
```

### What is in the dataset?

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

-   85 different types of candy.

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

-   38 fruity candy types.

## What is your favorite candy?

> Q3. What is your favorite candy in the dataset and what is it's winpercent value?

```{r}
candy["Junior Mints", ]$winpercent
```

-   My favorite candy in the dataset is Junior Mints. Its winpercent value is 57.21925

> Q4. What is the winpercent value for "Kit Kat"?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

**Side Note:** `skim()` can give a quick overview of a given dataset. Install the skimr package and try it on our candy data.

```{r}
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

-   winpercent is on a scale of 0-100 while the other variables are on a 0-1 scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

- A zero means there is no chocolate in the candy and a 1 means there is chocolate in the candy.


>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```
OR we can use ggplot: 
```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 10, col = "black", fill = "orange")
```


>Q9. Is the distribution of winpercent values symmetrical?

- The winpercent values are skewed toward the left. 

>Q10. Is the center of the distribution above or below 50%?

- The center is below 50%. 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate.inds <- as.logical(candy$chocolate)
chocolate.wins <- candy[chocolate.inds, ]$winpercent
mean(chocolate.wins)
```

```{r}
fruity.inds <- as.logical(candy$fruity)
fruity.wins <- candy[fruity.inds, ]$winpercent
mean(fruity.wins)
```

- Chocolate candy is ranked higher than fruity candy on average. 

>Q12. Is this difference statistically significant?

```{r}
t.test(chocolate.wins, fruity.wins)
```
- The p-value is less than 0.05, so the difference is statistically significant. 


## Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent), ])
```


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent), ])
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill = "plum", col = "orange")
```

### Time to add some useful color 

Let's set up a color vector that signifies candy type. 
Start by making a vector of all black values then we overwrite chocolate, brown, and red.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

Now let's try our barplot with the new colors 

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill = my_cols)
```

Save the plot as an image:
```{r}
ggsave("tmp.png")
```


>Q17. What is the worst ranked chocolate candy?

- Sixlets

>Q18. What is the best ranked fruity candy?

- Starburst


## Taking a Look at Pricepercent

Make a plot of winpercent vs pricepercent:
```{r}
library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 10)
```

Order the candy by pricepercent:
```{r}
ord <- order(candy$pricepercent, decreasing = T)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
candy[(order(candy$winpercent, decreasing = T)), c(11, 12)]
```


-  Reese's Miniatures has a pricepercent value of 0.279 and a winpercent value of 81.86626

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
head(candy[ord, c(11, 12)], n = 5)
```
- Nik L Nip is the least popular of the most expensive candies. 


## Exploring the Correlation Structure 

Let's see how the variables interact with one another using the corrplot package.

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

- fruity and chocolate. 

>Q23. Similarly, what two variables are most positively correlated?

- winpercent and chocolate. 


## Principal Component Analysis 

Apply PCA using the `prcom()` function to our candy dataset 

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```


Now we can plot our main PCA score plot of PC1 vs PC2 


```{r}
plot(pca$x[, 1], pca$x[, 2])
```

Change the plotting character and add some color: 

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

We can make a new data.frame that contains our PCA results with the rest of the candy data: 

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

The data.frame can be used to make a nicer plot using ggplot2: 

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

Use ggrepel to add labels:

```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```


Make a barplot of our findings: 

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

- fruity, hard, and pluribus. 






















